Bootstrapping Multi-atlas Hippocampal Segmentation with \mb

### 2013-08-18 paper.Rnw
Completed responses to Aristotle's comments

To validate that \mb algorithm works as well as established automated methods,
\mb is applied to the ADNI1 dataset and the resulting segmentations are
compared to those produced by FreeSurfer, FSL, MAPER, and by expert manual
segmentation.

\paragraph{ADNI1 dataset revisited} 
All images $1919$ images from the {\em \adnidataset} standardized dataset
described in Experiment 1 are used.  Clinical and demographic data for these
subjects are shown in Table \ref{tab:ADNI1-Dataset-Demographics}.

Based on the results from the ADNI1 Cross-Validation experiment, in this
experiment \mb was configured with a template library of 21 randomly chosen
subject images (7 from each disease class) and used majority vote label fusion.
The entire \adnidataset dataset was segmented by \mb and we now compare the
resulting volumes with those obtained by manual segmentation (SNT), and other
automated segmentation techniques (MAPER, FreeSurfer, and FSL).  Table
\ref{tab:ADNI1-1.5T-Complete-1Yr-Dataset-Method-Totals} shows the total count of
segmentations available, including a count of those which have failed a quality
control inspection. Quality control for FreeSurfer segmentations was provided along
with the segmentations from ADNI \cite{Hartig2010}. One of the authors (MP) 
performed visual quality inspection for MAGeT and FSL segmentations using similar 
quality control guidelines (if either hippocampus was under or over segmented
by greater than 10mm in three or more slices then the segmentation did not
pass).  Only those images which had quality segmentations from each method are
included in the following analysis (a total of
\Sexpr{length(means.complete$RID)} images).

\paragraph{Experiment details} 
\mb is configured with an atlas library composed of the Winterburn T1 atlases
(see Experiment 2). All images from the \adnidataset dataset are segmented. The
optimal size of template library, registration method, and label fusion method
found in Experiment 1 are used. The template library is composed of 21 randomly
chosen subject images (7 from each disease class).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%    RESULTS: Experiment 2: Hippocampal Subfield Cross-Validation     %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Experiment 2 Results: Winterburn Atlases Cross-Validation}
This experiment explores \mb segmentation of hippocampal subfields. To
achieve this, a leave-one-out validation is conducted in which lower-resolution
images ($0.9mm^3$ voxels) of each Winterburn atlas subject are segmented using 
the remaining Winterburn atlases.

@@

Based on the results from the ADNI1 Cross-Validation experiment, in this
experiment \mb was configured with a template library of 21 randomly chosen
subject images (7 from each disease class) and used majority vote label fusion.
The entire \adnidataset dataset was segmented by \mb and we now compare the
resulting volumes with those obtained by manual segmentation (SNT), and other
automated segmentation techniques (MAPER, FreeSurfer, and FSL).  Table
\ref{tab:ADNI1-1.5T-Complete-1Yr-Dataset-Method-Totals} shows the total count of
segmentations available, including a count of those which have failed a quality
control inspection. Quality control for FreeSurfer segmentations was provided along
with the segmentations from ADNI \cite{Hartig2010}. One of the authors (MP) 
performed visual quality inspection for MAGeT and FSL segmentations using similar 
quality control guidelines (if either hippocampus was under or over segmented
by greater than 10mm in three or more slices then the segmentation did not
pass).  Only those images which had quality segmentations from each method are
included in the following analysis (a total of
\Sexpr{length(means.complete$RID)} images).

We find a close relationship in total bilateral hippocampal volume between
all methods and manually segmented volumes (Figure \ref{fig:ADNI-volumes-plot}). 
Volumes are correlated with Pearson $r > 0.78$ for all methods across disease
categories.  Within disease categories (Figure \ref{fig:ADNI-volumes-boxplot}),
\mb is consistently well correlated to manual volumes (Pearson $r > 0.85$), but
appears to slightly over-estimate the volume of the AD hippocampus.

@@

Although, the goal of this manuscript was to not exhaustively test or validate
multiple different voting strategies in the context of our segmentation
algorithm, it is important to note that other strategies for voting are
available.  For example, other groups have used the STAPLE algorithm
\citep{Warfield2004} (or variants of the STAPLE algorithm \citep{Robitaille2012})
which weights each segmentation based upon its estimated
performance level with respect to the other available candidate segmentations.
Further, the sensitivity and specificity parameters could also be tuned to
potentially improve segmentation accuracy and reliability.  It is rather likely
that using more sophisticated voting methods would have a positive effect on the
overall segmentation performance.

To this end, more work is required in order to determine the source of the
slight decrease in segmentation performance when the number of templates are set
to an even number.  Our initial concern was that this dip in performance was a
by-product of the \mb algorithm itself.  However, we determined that
this pattern was also true in the analysis of multi-atlas segmentations we used
in our experiments.  We believe that our majority voting methodology is biased
towards labels with the lowest numeric values when breaking ties, thus causing
the slight bias observed when using an even number of templates. This is another
area where the voting scheme could be used to improve performance. However, it
is worth noting that this limitation was previously identified by
\citet{Heckemann2006a} and, subsequently, other groups have not even considered
the potential pitfalls of an even number of candidate labels (eg.
\citet{Leung2010}).

Another concern is the moderate-to-small improvement observed in \mb in
comparison to multi-atlas segmentation when using the same number of atlases.
The actual benefit in using \mb is consistency of the labelling regardless of
atlas or template choice. This is an important consideration that few have
touched on previously. The 10-fold Monte Carlo leave-one-out cross-validation that we present
in Experiment 1 is amongst one of the most stringent performed in the
multi-atlas/segmentation literature.  To the best of our knowledge, other groups
using ADNI data for validation do at most a single round of leave-one-out-validation
(Table \ref{tab:other-methods}), aside from \citep{Wang2011} who perform a 
similar validation as us. The thoroughness of our validation suggests that
our results are reflective of a true average over the choice of parameter
settings and are independent of atlas or template choice (provided the input
atlases are properly segmented).

@@

There are some important differences between our method
and these specific methods.  Others have reported the difficulty with
mis-registrations in candidate segmentation (i.e. segmentations generated that are
then input in the voxel-voting procedure \citep{Collins2010}). The work of 
\citet{Leung2010} tackles this problem by using an intensity threshold
that is estimated heuristically at the time of segmentation (this work also
reports some of the highest kappas for the segmentation of ADNI data).  While
this works for the ADNI dataset (which is partially homogenized with respect to
image acquisition and pre-processing), it is unclear if this type of heuristic
is applicable to other datasets.  In all cases, these methods require more
atlases than our implementation with the Winterburn atlases.
\citet{Lotjonen2010} achieve highly accurate segmentation but correct their
segmentations using classifications derived using an expectation maximization
framework.  In their initial work, \citet{Chupin2009} develop their
probabilistic methodology using a cohort of 8 healthy controls and 15 epilepsy
patients.  In subsequent methods, they retuned their methodology using the ADNI
sample with a hierarchical experimentation protocol that involved.  These
methods suggest that some post-processing of the final segmentations would
improve accuracy of the segmentation.  While that may be true, there is little
consensus regarding how to achieve this.

@@

In comparison to the methods that we compared and were available through the
ADNI database (FreeSurfer, MAPER) and the method that we initialized ourselves
(FSL-FIRST) we find extremely variable performance of all methods.  With the
exception of FSL all methods correlate well with the SNT volumes provided in
the ADNI database.  However, FreeSurfer and FIRST provide radically different
definitions of the size of the hippocampus in comparison to the other methods.
Further, when estimating bias of these methods relative to SNT hippocampal
volumes we see that large hippocampi are over estimated while small hippocampi
are under estimated.  By comparison, our method and MAPER are far more
conservative suggesting that these methods may be better suited for estimating
true-positives.  While this comparison with these methods using only volume,
more work is needed to better understand the differences between our method and
other methods.

Finally, we have also demonstrated that our algorithmic framework is appropriate
for the segmentation of hippocampal subfields in standard T1-weighted data.
This has started to become a burgeoning topic in the segmentation literature,
although very few methods are available for the segmentation of 3T data
\citep{Yushkevich2009,VanLeemput2009}. However, the FreeSurfer implementation of
subfield segmentation. While recent work demonstrates that subfield
segmentations can be used for classification of AD, MCI, and NC, there has been
no explicit validation of the methodology based on accuracy or precision.
Although the initial hippocampal subfield segmentations from the Yushkevich
group have been demonstrated to work on the ADNI population, there has also been
no validation of their work.  In addition, their work requires some manual
initialization to properly function.  Our work demonstrates that we can reliably
identify the CA1, subiculum, and CA4 dentate with only modest amounts of error.
The fact that CA2/CA3 and molecular layers cannot be reasonably identified
should not be surprising.  These are extremely thin and spatially convoluted
regions that originally required high-resolution MRI for identification.  It is
likely that they would be that extents of these regions are well below the
resolution offered by standard T1-weighted images.  In fact many manual
segmentation methodologies do not attempt to parse these regions either
\citep{Wisse2012, Mueller2009}

\subsection{Conclusion}
In conclusion, we have presented a flexible multi-atlas framework.  It has
considerable advantages over other methods as only a small set of atlases is
required to initialize the algorithm.  We demonstrate that our method works
robustly over hippocampal definitions, different populations, and different
acquisition types.  Finally, we also demonstrate that using this method that
accurate identification of the hippocampal subfields in also possible.

>>>>

To validate the \mb algorithm as respect to other established automated
methods, \mb is applied to large dataset from the ADNI project, and the
resulting segmentations are compared to those produced by FreeSurfer, FSL,
MAPER, and by expert semi-automated segmentation (SNT).

\paragraph{ADNI1 dataset revisited} 
The $1919$ images from the {\em \adnidataset} standardized dataset, described
in Experiment 1, were initially considered.  SNT and FreeSurfer hippocampal
volumes were provided by ADNI, along with FreeSurfer quality control data
(guidelines described in \citep{Hartig2010}). Clinical and demographic data for
the entire \adnidataset dataset are shown in Table
\ref{tab:ADNI1-Dataset-Demographics}.

\paragraph{Experiment details} 
\mb was configured with an atlas library composed of the five Winterburn T1
atlases (described in Experiment 2). The optimal size of template library,
registration method, and label fusion method found in Experiment 1 were used.
The FSL FIRST segmentation method was used  via the \verb+run_first_all+ script
according to the FIRST user guide
\footnote{http://fsl.fmrib.ox.ac.uk/fsl/fslwiki/FIRST/UserGuide}.  All images
from the \adnidataset dataset were segmented by both methods.

One author (MP) performed visual quality inspection for MAGeT and FSL
segmentations using similar quality control guidelines (if either hippocampus
was under or over segmented by greater than 10mm in three or more slices then
the segmentation did not pass).  Only images meeting the conditions of having
segmentations from all methods (SNT, MAPER, FreeSurfer, FSL, and \mb) and also
passing quality control inspection are included in the analysis.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%    RESULTS: Experiment 2: Hippocampal Subfield Cross-Validation     %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Experiment 2 Results: Winterburn Atlases Cross-Validation}
This experiment explores \mb segmentation of hippocampal subfields. To achieve
this, a leave-one-out validation was conducted in which lower-resolution images
($0.9mm^3$ voxels) of each Winterburn atlas subject were segmented using the
remaining Winterburn atlases.

@@

Based on the results from the ADNI1 Cross-Validation experiment, in this
experiment \mb was configured with a template library of 21 randomly chosen
subject images (7 from each disease class) and used majority vote label fusion.
The entire \adnidataset dataset was segmented by \mb, and the resulting volumes
compared with those obtained by expert semi-automated segmentation (SNT), and
three other automated segmentation techniques: MAPER, FreeSurfer, and FSL.
Table \ref{tab:ADNI1-1.5T-Complete-1Yr-Dataset-Method-Totals} shows the total
count of segmentations available, including a count of those which have failed
a quality control inspection.  A total of \Sexpr{length(means.complete$RID)}
images are included in the following analysis, having met quality criteria and
having segmentations from every method.

We found a close relationship in total bilateral hippocampal volume between
all methods and manually segmented volumes (Figure \ref{fig:ADNI-volumes-plot}). 
Volumes are correlated with Pearson $r > 0.78$ for all methods across disease
categories.  Within disease categories (Figure \ref{fig:ADNI-volumes-boxplot}),
\mb is consistently well correlated to manual volumes (Pearson $r > 0.85$), but
appears to slightly over-estimate the volume of the AD hippocampus.

@@

Although, the goal of this manuscript was to not exhaustively test or validate
multiple different voting strategies in the context of our segmentation
algorithm, it is important to note that other strategies for voting are
available.  For example, other groups have used the STAPLE algorithm
\citep{Warfield2004} (or variants of the STAPLE algorithm
\citep{Robitaille2012}) which weights each segmentation based upon its
estimated performance level with respect to the other available candidate
segmentations.  Further, the sensitivity and specificity parameters could also
be tuned to potentially improve segmentation accuracy and reliability.  It is
likely that using more sophisticated voting methods would have a positive
effect on the overall segmentation performance, as demonstrated by the STAPLE
algorithm.

To this end, more work is required in order to determine the source of the
slight decrease in segmentation performance when the number of templates are
set to an even number.  Our initial concern was that this dip in performance
was a by-product of the \mb algorithm itself.  However, this this pattern is
also found in the results of the multi-atlas segmentations we used in our
experiments.  We believe that our majority voting methodology is biased towards
labels with the lowest numeric values when breaking ties (by way of the
implementation of the \verb+mode+ function used to determine majority), thus
causing the slight bias observed when using an even number of templates. This
is another area where the voting scheme could be used to improve performance.
However, it is worth noting that this limitation was previously identified by
\citet{Heckemann2006a} and, subsequently, other groups have not even considered
the potential pitfalls of an even number of candidate labels (eg.
\citet{Leung2010}).

Another concern is the moderate-to-small improvement observed in \mb in
comparison to multi-atlas segmentation when using the same number of atlases.
The actual benefit in using \mb is consistency of the labelling regardless of
atlas or template choice. This is an important consideration that few have
touched on previously. The 10-fold Monte Carlo leave-one-out cross-validation that we present
in Experiment 1 is amongst one of the most stringent performed in the
multi-atlas/segmentation literature. To the best of our knowledge, other groups
using ADNI data for validation do at most a single round of
leave-one-out-validation (Table \ref{tab:other-methods}), aside from
\citep{Wang2011} who perform a similar validation as us). The thoroughness of
our validation suggests that our results are reflective of a true average over
the choice of parameter settings and are independent of atlas or template
choice (provided the input atlases are properly segmented.

@@

There are some important differences between our method
and these specific methods.  Others have reported the difficulty with
mis-registrations in candidate segmentation (i.e. segmentations generated that are
then input in the voxel-voting procedure \citep{Collins2010}). The work of 
\citet{Leung2010} tackles this problem by using an intensity threshold
that is estimated heuristically at the time of segmentation (this work also
reports some of the highest kappas for the segmentation of ADNI data).  While
this works for the ADNI dataset (which is partially homogenized with respect to
image acquisition and pre-processing), it is unclear if this type of heuristic
is applicable to other datasets.  In all cases, these methods require more
atlases than our implementation with the Winterburn atlases.
\citet{Lotjonen2010} achieve highly accurate segmentation but correct their
segmentations using classifications derived using an expectation maximization
framework. In their initial work, \citet{Chupin2009} develop their
probabilistic methodology using a cohort of 8 healthy controls and 15 epilepsy
patients, and then use this method to segment an ADNI sample, with a
hierarchical experimentation protocol that involved \todo{complete this
sentence}.  These methods suggest that some post-processing of the final
segmentations would improve accuracy of the segmentation. While that may be
true, there is little consensus regarding how to achieve this.

@@

In comparison to the methods that we compared and were available through the
ADNI database (FreeSurfer, MAPER) and the method that we initialized ourselves
(FSL-FIRST) we find extremely variable performance of all methods.  With the
exception of FSL all methods correlate well with the SNT volumes provided in
the ADNI database.  However, FreeSurfer and FIRST provide radically different
definitions of the size of the hippocampus in comparison to the other methods.
Further, when estimating bias of these methods relative to SNT hippocampal
volumes we see that large hippocampi are over estimated while small hippocampi
are under estimated. By comparison, \mb and MAPER are far more conservative in
volume estimation, suggesting these methods may be better suited for estimating
true-positives. However, in this analysis we have only compared methods by
total hippocampal volume, and so more work is needed to understand the full
extent to which these methods differ.

Finally, we have also demonstrated that our algorithmic framework is
appropriate for the segmentation of hippocampal subfields in standard
T1-weighted data.  Subfield segmentation is a burgeoning topic in the
literature although very few automated methods are available for the
segmentation of 3T data \citep{Yushkevich2009,VanLeemput2009}. However, the
FreeSurfer implementation of subfield segmentation \todo{complete}. While
recent work demonstrates that subfield segmentations can be used for
classification of AD, MCI, and NC, there has been no explicit validation of the
methodology based on accuracy or precision.  Although the initial hippocampal
subfield segmentations from the Yushkevich group have been demonstrated to work
on the ADNI population, there has also been no validation of their work.  In
addition, their work requires some manual initialization to properly function.
Our work demonstrates that we can reliably identify the CA1, subiculum, and CA4
dentate with only modest amounts of error.  The fact that CA2/CA3 and molecular
layers cannot be reasonably identified should not be surprising as these are
extremely thin and spatially convoluted regions that originally required
high-resolution MRI for identification.  It is likely that extents of these
regions are well below the resolution offered by standard T1-weighted images
and, in fact, many manual segmentation methodologies do not attempt to parse
these regions either \citep{Wisse2012, Mueller2009}.

In conclusion, we have presented a flexible multi-atlas framework.  It has
considerable advantages over other methods as only a small set of atlases is
required to initialize the algorithm.  We demonstrate that our method works
robustly over hippocampal definitions, different populations, and different
acquisition types.  Finally, we also demonstrate that using this method that
accurate identification of the hippocampal subfields in also possible.
